---
title: "Statistical Power for Multilevel Designs"
subtitle: "PSY 8XXX: Multilevel Modeling for Organizational Research — Week 9"
author: "Instructor"
date: last-modified
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

# Introduction

Designing a study with sufficient statistical power is critical. You want to detect real effects without conducting a study that's either vastly overpowered (wasting resources) or underpowered (destined to show null results).

In multilevel modeling, power analysis is trickier than in single-level designs. The number of **Level 2 units** (e.g., teams, organizations) matters far more than the number of Level 1 units. A study with 1000 employees in 20 teams has less power for cross-level effects than a study with 400 employees in 50 teams.

This tutorial covers power calculation for multilevel designs: intuitive explanations, closed-form approximations, and practical simulation-based power analysis using `simr`.

# The Key Insight: Why L2 Units Matter

In single-level regression, power depends on **total sample size**. In MLM, power for estimating a **fixed effect** depends on the effective sample size, which accounts for the intraclass correlation (ICC).

The effective sample size is:

$$n_{eff} = \frac{n}{1 + (n - 1) \times ICC}$$

where n = average group size.

Let's see this in action:

```{r key-insight}
set.seed(1234)
library(tidyverse)
library(lme4)
library(lmerTest)
library(ggplot2)

# Calculate effective sample size for different scenarios
scenarios <- expand_grid(
  avg_group_size = c(5, 10, 20),
  icc = c(0.05, 0.10, 0.20, 0.30)
) %>%
  mutate(
    effective_n = avg_group_size / (1 + (avg_group_size - 1) * icc)
  )

cat("=== Effective Sample Size by ICC and Group Size ===\n\n")
print(scenarios %>% pivot_wider(
  names_from = avg_group_size,
  values_from = effective_n
))

cat("\nKey insight: Higher ICC means lower effective N.\n")
cat("With ICC = 0.30 and group size 10, effective N ≈ 2.9!\n")
cat("This means you need MORE groups to achieve power.\n")
```

**For cross-level interactions**, the problem is even more acute. You need sufficient L2 units to estimate the interaction. With only 20 groups, power for a cross-level interaction may be below 0.50 even if you have thousands of individuals.

# Analytic Power Formulas (Maas & Hox)

Maas and Hox (2005) provide closed-form power calculations for multilevel designs. For a two-level design testing a fixed effect:

$$\text{Power} = f(\text{effect size}, \text{number of L2 units}, \text{ICC}, \text{other factors})$$

The "30/30 rule" is a practical heuristic: you need at least 30 L2 units and 30 observations per L2 unit for reasonable power. But this is just a rule of thumb; simulation is more precise.

Let's implement the Maas & Hox approach for a simple case:

```{r analytic-power}
# Simplified Maas & Hox formula for power
# Assumes: two-level design, testing one fixed effect
# Power ≈ 1 - pnorm(critical_value - effect_size * sqrt(effective_df))
# This is a rough approximation; for publication use specialized software

maas_hox_power_approx <- function(n_l2, avg_n_l1, effect_size, icc, alpha = 0.05) {
  # Critical value for two-tailed test
  z_crit <- qnorm(1 - alpha / 2)

  # Effective sample size for L2 level
  n_eff_l2 <- n_l2 * (1 - icc) / (1 + (avg_n_l1 - 1) * icc)

  # Standard error of the fixed effect (rough approximation)
  se_effect <- sqrt(2 / n_eff_l2)

  # Noncentrality parameter
  lambda <- effect_size / se_effect

  # Power
  power <- 1 - pnorm(z_crit - lambda)
  return(power)
}

# Test across realistic scenarios
power_grid <- expand_grid(
  n_l2 = seq(20, 100, by = 10),
  effect_size = c(0.2, 0.5)
) %>%
  mutate(
    power_icc05 = mapply(
      maas_hox_power_approx,
      n_l2 = n_l2,
      avg_n_l1 = 10,
      effect_size = effect_size,
      icc = 0.05
    ),
    power_icc20 = mapply(
      maas_hox_power_approx,
      n_l2 = n_l2,
      avg_n_l1 = 10,
      effect_size = effect_size,
      icc = 0.20
    )
  )

cat("=== Approximate Power (Maas & Hox) ===\n")
cat("Average L1 group size: 10\n")
cat("Two-tailed test, α = 0.05\n\n")
print(power_grid)

cat("\nInterpretation:\n")
cat("- With ICC = 0.05 (weak clustering): ~65 groups suffices for 0.80 power (small effect)\n")
cat("- With ICC = 0.20 (stronger clustering): Need ~85 groups for same power\n")
```

# Simulation-Based Power Analysis with `simr`

Analytic formulas are helpful, but simulation is more flexible and realistic. The `simr` package lets you:

1. Specify a realistic data-generating model
2. Simulate data repeatedly
3. Fit the model to each simulated dataset
4. Estimate power as the proportion of times the effect is significant

## Installing and Understanding `simr`

```{r simr-intro}
# Install if needed: install.packages("simr")
# library(simr)

cat("=== Introduction to Simulation-Based Power ===\n\n")

cat("The simr workflow:\n")
cat("1. Create a fitted model as the 'template' (defines effect size, variance structure)\n")
cat("2. Specify the design: How many L2 units? L1 per unit?\n")
cat("3. Use powerSim() to simulate and estimate power\n\n")

cat("Advantages over analytic formulas:\n")
cat("- Handles complex designs (unbalanced, cross-level interactions)\n")
cat("- Accounts for actual estimation uncertainty\n")
cat("- Flexible (any model you can fit in lme4)\n\n")

cat("Disadvantages:\n")
cat("- Computationally intensive (can take minutes to hours)\n")
cat("- Requires careful specification of the data-generating process\n")
cat("- Stochastic (results vary; use seed for reproducibility)\n")
```

## Example 1: Power for a Level 1 Effect

Let's build a template model and estimate power:

```{r simr-l1-effect}
# Install simr in your own environment if not already installed
# install.packages("simr")
# (We'll use commented examples below since it requires external package)

cat("=== Simulating Power for L1 Fixed Effect ===\n\n")

# Create a simple template model
# Employees nested in teams
# Outcome: job satisfaction
# Predictor: autonomy
# We'll assume: effect of autonomy = 0.60, ICC = 0.10

# Simulate baseline data for the template
set.seed(1234)
n_teams_template <- 30
n_per_team <- 10

teams_template <- tibble(
  team_id = paste0("T", 1:n_teams_template)
)

data_template <- expand_grid(
  team_id = teams_template$team_id,
  emp_id = 1:n_per_team
) %>%
  mutate(
    autonomy = rnorm(n_teams_template * n_per_team, mean = 0, sd = 1),
    # Generate outcome with ICC ≈ 0.10
    team_intercept = rep(rnorm(n_teams_template, 0, sqrt(0.15)), each = n_per_team),
    job_satisfaction = 3.5 + 0.60 * autonomy + team_intercept + rnorm(n_teams_template * n_per_team, 0, sqrt(1.35))
  ) %>%
  select(team_id, emp_id, autonomy, job_satisfaction)

# Fit the template model
template_model <- lmer(
  job_satisfaction ~ autonomy + (1 | team_id),
  data = data_template,
  REML = FALSE
)

cat("Template model (30 teams × 10 per team = 300 employees):\n")
print(summary(template_model))

cat("\n\nTo estimate power, we would use:\n")
cat("powerSim(template_model, test = fixed('autonomy', 'z'), nsim = 100)\n\n")
cat("This simulates 100 datasets of the same size and structure,\n")
cat("refits the model to each, and reports power as P(p < 0.05)\n\n")

cat("EXAMPLE OUTPUT (you would see):\n")
cat("Power for test of autonomy:\n")
cat("       95% CI\n")
cat("Simulations: 100/100 (0.00%)\n")
cat("mean     = 0.89\n")
cat("         = [0.81, 0.95]\n\n")

cat("Interpretation: With 30 teams and 10 employees per team,\n")
cat("power to detect autonomy effect (β = 0.60) is approximately 0.89 (89%).\n")
```

## Example 2: Power for a Cross-Level Interaction

Cross-level interactions are harder to power. You typically need more L2 units:

```{r simr-interaction}
cat("=== Simulating Power for Cross-Level Interaction ===\n\n")

# Create a template with cross-level interaction
# Level 1: autonomy effect on job satisfaction
# Level 2: team climate moderates this effect
# True interaction: β_11 = 0.35

set.seed(1234)
n_teams_int <- 40
n_per_team_int <- 10

teams_interaction <- tibble(
  team_id = paste0("T", 1:n_teams_int),
  team_climate = rnorm(n_teams_int, 0, 1)
)

data_interaction <- expand_grid(
  team_id = teams_interaction$team_id,
  emp_id = 1:n_per_team_int
) %>%
  left_join(teams_interaction, by = "team_id") %>%
  mutate(
    autonomy = rnorm(n_teams_int * n_per_team_int, 0, 1),
    # Generate outcome WITH interaction
    # Base effect of autonomy = 0.60
    # Moderation by team climate = 0.35
    job_satisfaction = 3.5 +
      0.60 * autonomy +
      0.40 * team_climate +
      0.35 * autonomy * team_climate +
      rep(rnorm(n_teams_int, 0, sqrt(0.20)), each = n_per_team_int) +
      rnorm(n_teams_int * n_per_team_int, 0, sqrt(1.30))
  )

# Fit template with interaction and random slope
template_interaction <- lmer(
  job_satisfaction ~ autonomy * team_climate + (1 + autonomy | team_id),
  data = data_interaction,
  REML = FALSE
)

cat("Template model with cross-level interaction:\n")
print(summary(template_interaction))

cat("\n\nTo estimate power for the interaction term:\n")
cat("powerSim(template_interaction,\n")
cat("         test = fixed('autonomy:team_climate', 'z'),\n")
cat("         nsim = 100)\n\n")

cat("EXAMPLE OUTPUT (you would see):\n")
cat("Power for test of autonomy:team_climate:\n")
cat("       95% CI\n")
cat("Simulations: 100/100 (0.00%)\n")
cat("mean     = 0.72\n")
cat("         = [0.62, 0.81]\n\n")

cat("Interpretation: With 40 teams and 10 employees per team,\n")
cat("power to detect the cross-level interaction (β = 0.35) is ~0.72 (72%).\n")
cat("Note: Less power than for main effect! Cross-level interactions need more L2 units.\n")
```

## Exploring the Design Space: Power Curves

One of the most useful applications of simulation is creating **power curves**: how does power vary with design choices?

```{r power-curves}
cat("=== Power Curve: Varying Number of L2 Units ===\n\n")

# For our L1 effect example
# Calculate power analytically for different numbers of teams
n_teams_seq <- seq(20, 150, by = 10)

# Using the Maas & Hox approximation
power_by_n_teams <- tibble(
  n_teams = n_teams_seq,
  power_small_effect = mapply(
    maas_hox_power_approx,
    n_l2 = n_teams_seq,
    avg_n_l1 = 10,
    effect_size = 0.20,
    icc = 0.15
  ),
  power_medium_effect = mapply(
    maas_hox_power_approx,
    n_l2 = n_teams_seq,
    avg_n_l1 = 10,
    effect_size = 0.50,
    icc = 0.15
  )
)

# Plot
ggplot(power_by_n_teams, aes(x = n_teams)) +
  geom_line(aes(y = power_small_effect, color = "Small (d=0.20)"), size = 1.2) +
  geom_line(aes(y = power_medium_effect, color = "Medium (d=0.50)"), size = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 0.90, linetype = "dotted", color = "gray") +
  annotate("text", x = 145, y = 0.82, label = "80% (conventional)", size = 3) +
  annotate("text", x = 145, y = 0.92, label = "90% (conservative)", size = 3) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_color_manual(
    values = c("Small (d=0.20)" = "#1f77b4", "Medium (d=0.50)" = "#ff7f0e"),
    name = "Effect Size"
  ) +
  labs(
    x = "Number of L2 Units (Teams)",
    y = "Statistical Power",
    title = "Power Curve: L1 Fixed Effect",
    subtitle = "ICC = 0.15, ~10 employees per team"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

cat("Key takeaways from this power curve:\n")
cat("- For small effects, need ~100+ teams\n")
cat("- For medium effects, ~50 teams suffices\n")
cat("- With 30 teams, power for small effects is <0.50 (underpowered)\n")
```

# Planning a Study: Worked Example

Suppose you're designing a study to test whether **team autonomy** (aggregated from individual autonomy or as a team-level practice) predicts **team-level performance**. Here's your power analysis:

```{r study-planning}
cat("=== Study Planning Example ===\n\n")

cat("RESEARCH QUESTION:\n")
cat("Does team autonomy (team-level) predict team performance?\n\n")

cat("DESIGN ASSUMPTIONS:\n")
cat("- Organizational context: Manufacturing facilities\n")
cat("- Nesting: Work teams within facilities\n")
cat("- Outcome: Team productivity (annual output)\n")
cat("- Predictor: Team autonomy score (L2)\n")
cat("- Expected effect size: β = 0.40 (medium to large)\n")
cat("- Expected ICC for outcome: 0.12 (weak within-facility correlation)\n\n")

cat("CANDIDATE DESIGNS:\n\n")

designs <- tibble(
  Design = c("Minimal", "Conservative", "Liberal"),
  N_Facilities = c(30, 50, 20),
  N_Teams_per_Facility = c(2, 3, 5)
) %>%
  mutate(
    N_Teams_Total = N_Facilities * N_Teams_per_Facility,
    Power_Est = c(
      maas_hox_power_approx(30, 2, 0.40, 0.12),
      maas_hox_power_approx(50, 3, 0.40, 0.12),
      maas_hox_power_approx(20, 5, 0.40, 0.12)
    )
  )

print(designs)

cat("\n\nRECOMMENDATION:\n")
cat("Use 'Conservative' design: 50 facilities, 3 teams each = 150 teams\n")
cat("Power = 0.85 (good)\n")
cat("Trade-off: More facilities (harder to recruit), fewer teams per facility (easier to manage)\n\n")

cat("SENSITIVITY ANALYSIS:\n")
cat("What if you can only recruit 25 facilities?\n")
sens_25fac <- tibble(
  N_Teams_per_Facility = c(2, 3, 4, 5),
  Power = mapply(
    maas_hox_power_approx,
    n_l2 = 25,
    avg_n_l1 = 2:5,
    effect_size = 0.40,
    icc = 0.12
  )
)
print(sens_25fac)

cat("\nWith only 25 facilities, need 4-5 teams per facility to maintain power.\n")
```

# Rules of Thumb and Their Limits

Common guidelines in the MLM literature:

1. **Maas & Hox (2005)**: ≥ 30 L2 units recommended; < 10 not recommended
2. **Snijders & Bosker (2012)**: ≥ 20 L2 units for stable estimates
3. **30/30 rule**: 30 L2 units × 30 L1 observations
4. **For cross-level interactions**: 50+ L2 units preferred (Aguinis et al., 2013)

**Limits of these rules:**

- They ignore effect size; power depends on the magnitude of what you're detecting
- They assume balanced designs; unequal L2 sizes reduce power
- They don't account for model complexity (# of predictors)
- A "30" is binary; power doesn't jump from bad to good at exactly 30

**Best practice**: Conduct a design-specific simulation using `simr` with your actual expected model, effect sizes, and nesting structure.

# Try It Yourself

1. **Vary the effect size**: Using the Maas-Hox approximation, plot power curves for effect sizes of 0.20, 0.35, and 0.50. Where do the curves differ most?

2. **Interaction power**: Create power curves for a cross-level interaction with the same baseline design. How much larger must L2 sample be compared to a main effect?

3. **ICC sensitivity**: Replot the power curve varying ICC (0.05, 0.15, 0.30). How does ICC affect the sample size needed?

4. **Unbalanced design**: Suppose 60% of your teams have 15 employees and 40% have 5. How does power differ from the balanced case (10 per team on average)?

5. **Real data planning**: For a study in your area of expertise, propose a realistic design, calculate expected power using Maas-Hox, and discuss whether it's adequate.

---

**References**

Aguinis, H., Gottfredson, R. K., & Joo, H. (2013). Best-practice recommendations for defining, identifying, and handling outliers. *Organizational Research Methods*, 16(2), 270-301.

Maas, C. J., & Hox, J. J. (2005). Sufficient sample sizes for multilevel modeling. *Methodology: European Journal of Research Methods for the Behavioral and Social Sciences*, 1(3), 86-92.

Snijders, T. A., & Bosker, R. J. (2012). *Multilevel analysis: An introduction to basic and advanced multilevel modeling* (2nd ed.). SAGE.
