---
title: "Multilevel Mediation: Within and Between Effects"
subtitle: "PSY 8XXX: Multilevel Modeling for Organizational Research — Week 10"
author: "Instructor"
date: last-modified
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

# Introduction

In organizational psychology, we often hypothesize **mediation**: X affects outcome Y through a mechanism (mediator M). For example:

- Does **team climate** affect **team performance** via increased **member autonomy**?
- Does **leader support** affect **job satisfaction** through **psychological safety**?

When data are nested (employees within teams), a naive single-level mediation analysis **conflates** within-team and between-team effects. This tutorial shows you how to decompose mediation properly.

# The Problem: Why Naive Mediation Fails with Nested Data

Suppose we run a standard mediation analysis ignoring nesting:

$$M_{ij} = i_M + c_1 X_{ij} + e_{M,ij}$$
$$Y_{ij} = i_Y + c_2 X_{ij} + c_3 M_{ij} + e_{Y,ij}$$

Then test: Does the effect of X on Y decrease when we include M?

**The problem**: This mixes effects at two levels:
- X might affect M at the within-team level (how individual differences in X drive M)
- X might affect M at the team level (how team-level variation in X correlates with team M)
- Same for the M → Y relationship

When you pool these, you get a **biased** mediation estimate.

## Illustration

```{r setup}
set.seed(1234)
library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(ggplot2)

# Simulate data with DIFFERENT within and between mediation effects

n_teams <- 30
n_per_team <- 15
n_total <- n_teams * n_per_team

# Create teams
teams <- tibble(
  team_id = 1:n_teams,
  # Team-level X: team autonomy practices
  X_team = rnorm(n_teams, mean = 5, sd = 1.5),
  # Introduce a between-level mediation mechanism
  # team_autonomy → team_psychological_safety → team_performance
  M_team = 3 + 0.80 * X_team + rnorm(n_teams, 0, 1),
  Y_team = 50 + 0.60 * M_team + rnorm(n_teams, 0, 5)
)

# Create employees within teams
employees <- expand_grid(
  team_id = 1:n_teams,
  emp_within_team = 1:n_per_team
) %>%
  left_join(teams, by = "team_id") %>%
  mutate(
    employee_id = row_number(),
    # Individual-level X: individual autonomy (varies within and between teams)
    X_ind = X_team + rnorm(n_total, 0, 1.2),
    # Within-team mediation: autonomy → psychological_safety → satisfaction
    # Smaller effect at L1
    M_ind = 3 + 0.35 * (X_ind - X_team) + M_team + rnorm(n_total, 0, 1),
    # Outcome: job satisfaction (L1)
    Y_ind = 50 + 0.40 * (M_ind - M_team) + Y_team + rnorm(n_total, 0, 3),
    # Standardize for easier interpretation
    X_ind_c = scale(X_ind)[, 1],
    M_ind_c = scale(M_ind)[, 1],
    Y_ind_c = scale(Y_ind)[, 1]
  )

cat("=== Multilevel Data Structure ===\n")
cat("Teams: ", n_distinct(employees$team_id), "\n")
cat("Employees per team: ", n_per_team, "\n")
cat("Total observations: ", nrow(employees), "\n\n")

# Show a sample
print(employees %>% select(employee_id, team_id, X_ind, M_ind, Y_ind) %>% head(10))
```

Now let's show the problem with naive mediation:

```{r naive-mediation-problem}
# Naive approach: Ignore nesting, run standard mediation
cat("=== NAIVE MEDIATION (Ignoring Nesting) ===\n\n")

# Total effect: X → Y
model_total <- lm(Y_ind_c ~ X_ind_c, data = employees)
total_effect <- coef(model_total)["X_ind_c"]
cat("Total effect (X → Y): ", round(total_effect, 3), "\n\n")

# Direct effect (controlling for M): X → Y | M
model_direct <- lm(Y_ind_c ~ X_ind_c + M_ind_c, data = employees)
direct_effect <- coef(model_direct)["X_ind_c"]
cat("Direct effect (X → Y | M): ", round(direct_effect, 3), "\n\n")

# Indirect effect (simple subtraction)
indirect_naive <- total_effect - direct_effect
cat("Indirect effect (naive): ", round(indirect_naive, 3), "\n")
cat("Interpretation: ", round(indirect_naive / total_effect * 100, 1), "% of X's effect goes through M\n\n")

# The problem: This estimate is BIASED because it mixes levels
cat("PROBLEM: This estimate mixes within-team and between-team effects!\n")
cat("It doesn't tell us whether M mediates at L1, L2, or both.\n")
```

# Types of Multilevel Mediation

There are different patterns depending on what level each variable lives at:

- **1-1-1 Mediation**: L1 predictor → L1 mediator → L1 outcome (within teams)
- **2-1-1 Mediation**: L2 predictor → L1 mediator → L1 outcome (between-team → individual)
- **2-2-1 Mediation**: L2 predictor → L2 mediator → L1 outcome
- **1-1-2 Mediation**: L1 predictor → L1 mediator → L2 outcome

In organizational research, **2-1-1** is common: team climate (L2) affects individual behavior (M, L1) which affects individual outcome (Y, L1).

Our example has mixed patterns: some within-team mediation (1-1-1) and some between-team (2-2-1).

# The Solution: Decompose Within and Between Effects

The solution, developed by **Zhang, Zyphur, & Preacher (2009)**, is to:

1. **Center L1 variables** at the team mean
2. **Separate within-team** (deviation from team mean) and **between-team** (team mean) effects
3. Model these explicitly
4. Compute indirect effects at each level separately

Here's the approach:

```{r decomposition-approach}
cat("=== DECOMPOSING WITHIN AND BETWEEN EFFECTS ===\n\n")

# Step 1: Compute team-level aggregates and within-team deviations
employees <- employees %>%
  group_by(team_id) %>%
  mutate(
    # Team means
    X_team_mean = mean(X_ind),
    M_team_mean = mean(M_ind),
    Y_team_mean = mean(Y_ind),
    # Within-team deviations (person - team mean)
    X_within = X_ind - X_team_mean,
    M_within = M_ind - M_team_mean,
    Y_within = Y_ind - Y_team_mean
  ) %>%
  ungroup() %>%
  mutate(
    # Standardize for interpretation
    X_within_c = scale(X_within)[, 1],
    X_team_mean_c = scale(X_team_mean)[, 1],
    M_within_c = scale(M_within)[, 1],
    M_team_mean_c = scale(M_team_mean)[, 1],
    Y_within_c = scale(Y_within)[, 1]
  )

cat("Decomposed variables:\n")
cat("- X_within: How much person i deviates from their team's mean on X\n")
cat("- X_team_mean: Their team's average X\n")
cat("- Similarly for M and Y\n\n")

cat("Now we model within and between effects separately.\n")
```

## Implementing 1-1-1 Mediation (Within-Team)

Let's first test whether M mediates the X → Y relationship **within teams** (controlling for between-team variation):

```{r mediation-1-1-1}
cat("=== 1-1-1 MEDIATION: WITHIN-TEAM LEVEL ===\n\n")

cat("Model A: X_within → M_within\n")
model_xm_within <- lmer(
  M_within ~ X_within + (1 | team_id),
  data = employees,
  REML = FALSE
)
a_within <- fixef(model_xm_within)["X_within"]
cat("Effect of autonomy on psychological safety (within): a = ", round(a_within, 3), "\n\n")

cat("Model B: X_within, M_within → Y_within\n")
model_my_within <- lmer(
  Y_within ~ X_within + M_within + (1 | team_id),
  data = employees,
  REML = FALSE
)
c_prime_within <- fixef(model_my_within)["X_within"]
b_within <- fixef(model_my_within)["M_within"]
cat("Direct effect of autonomy on satisfaction (controlling M): c' = ", round(c_prime_within, 3), "\n")
cat("Effect of psychological safety on satisfaction: b = ", round(b_within, 3), "\n\n")

# Indirect effect
indirect_within <- a_within * b_within
cat("INDIRECT EFFECT (within-team): a × b = ", round(a_within, 3), " × ", round(b_within, 3),
    " = ", round(indirect_within, 3), "\n\n")

cat("Interpretation:\n")
cat("Within a team, for every 1 SD increase in individual autonomy,\n")
cat("psychological safety increases by ", round(a_within, 3), " SD,\n")
cat("which in turn increases satisfaction by ", round(indirect_within, 3), " SD.\n")
```

## Implementing 2-1-1 Mediation (Between-Team to Individual)

Now test whether team-level X affects individual outcomes via an individual-level mediator:

```{r mediation-2-1-1}
cat("=== 2-1-1 MEDIATION: TEAM CLIMATE → INDIVIDUAL SAFETY → SATISFACTION ===\n\n")

cat("Model A: Team-level X → Individual-level M\n")
cat("(Regress M on team mean of X, accounting for individual variation)\n\n")

model_xm_2_1 <- lmer(
  M_within ~ 1 + X_team_mean + (1 | team_id),
  data = employees,
  REML = FALSE
)
a_2_1 <- fixef(model_xm_2_1)["X_team_mean"]
cat("Effect of team autonomy on individual psychological safety: a = ", round(a_2_1, 3), "\n\n")

cat("Model B: Individual M → Individual Y (with L2 predictor)\n")
cat("(Test whether individual M mediates effect of team-level X on Y)\n\n")

model_my_2_1 <- lmer(
  Y_within ~ X_team_mean + M_within + (1 | team_id),
  data = employees,
  REML = FALSE
)
c_prime_2_1 <- fixef(model_my_2_1)["X_team_mean"]
b_2_1 <- fixef(model_my_2_1)["M_within"]
cat("Direct effect of team autonomy on satisfaction: c' = ", round(c_prime_2_1, 3), "\n")
cat("Effect of individual safety on satisfaction: b = ", round(b_2_1, 3), "\n\n")

# Indirect effect
indirect_2_1 <- a_2_1 * b_2_1
cat("INDIRECT EFFECT (2-1-1): a × b = ", round(a_2_1, 3), " × ", round(b_2_1, 3),
    " = ", round(indirect_2_1, 3), "\n\n")

cat("Interpretation:\n")
cat("Teams with higher autonomy practices have more psychologically safe members,\n")
cat("who report higher satisfaction.\n")
cat("The cross-level indirect effect: ", round(indirect_2_1, 3), " (standardized)\n")
```

# Bootstrapping Confidence Intervals for Indirect Effects

Point estimates aren't enough; we need confidence intervals. The **percentile bootstrap** is the standard approach:

```{r bootstrapping-indirect-effects}
cat("=== BOOTSTRAPPING INDIRECT EFFECTS ===\n\n")

cat("Workflow:\n")
cat("1. Sample observations with replacement\n")
cat("2. Refit mediation models on each bootstrap sample\n")
cat("3. Compute indirect effect each time\n")
cat("4. Use 2.5th and 97.5th percentiles as 95% CI\n\n")

# Bootstrap function for 1-1-1 mediation
bootstrap_indirect_1_1_1 <- function(data, n_boot = 1000) {
  indirect_effects <- numeric(n_boot)

  for (b in 1:n_boot) {
    # Resample with replacement at team level, preserving nesting
    team_ids_boot <- sample(unique(data$team_id), replace = TRUE)
    data_boot <- map_df(team_ids_boot, ~ data %>% filter(team_id == .x))

    # Refit models
    m_xm <- lmer(M_within ~ X_within + (1 | team_id), data = data_boot, REML = FALSE)
    m_my <- lmer(Y_within ~ X_within + M_within + (1 | team_id), data = data_boot, REML = FALSE)

    # Compute indirect effect
    a <- fixef(m_xm)["X_within"]
    b <- fixef(m_my)["M_within"]
    indirect_effects[b] <- a * b
  }

  return(indirect_effects)
}

# Run bootstrap
set.seed(1234)
cat("Running 1000 bootstrap samples (this takes a moment)...\n")
boot_effects_1_1_1 <- bootstrap_indirect_1_1_1(employees, n_boot = 1000)

# Compute CI
ci_lower <- quantile(boot_effects_1_1_1, 0.025)
ci_upper <- quantile(boot_effects_1_1_1, 0.975)
mean_effect <- mean(boot_effects_1_1_1)

cat("=== BOOTSTRAP RESULTS: 1-1-1 Mediation ===\n\n")
cat("Indirect effect: ", round(mean_effect, 3), "\n")
cat("95% Confidence Interval: [", round(ci_lower, 3), ", ", round(ci_upper, 3), "]\n\n")

if (ci_lower > 0 | ci_upper < 0) {
  cat("✓ Significant (CI does not include zero)\n")
} else {
  cat("✗ Not significant at 0.05 level\n")
}

# Visualize bootstrap distribution
boot_df <- tibble(effect = boot_effects_1_1_1)

ggplot(boot_df, aes(x = effect)) +
  geom_histogram(bins = 40, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = ci_lower, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = ci_upper, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean_effect, color = "blue", size = 1) +
  annotate("text", x = ci_lower, y = 100, label = "2.5%", angle = 90, hjust = 1.2) +
  annotate("text", x = ci_upper, y = 100, label = "97.5%", angle = 90, hjust = -0.2) +
  labs(
    x = "Indirect Effect (a × b)",
    y = "Frequency",
    title = "Bootstrap Distribution of 1-1-1 Indirect Effect",
    subtitle = "Red dashed lines show 95% CI"
  ) +
  theme_minimal()
```

# Presenting Multilevel Mediation Results

A publication-ready table for mediation results:

```{r results-table}
cat("=== PUBLICATION TABLE: Multilevel Mediation Results ===\n\n")

results_table <- tibble(
  Pathway = c(
    "X → M (within)",
    "M → Y (within)",
    "Indirect (within)",
    "X → M (team level)",
    "M → Y (within | team X)",
    "Indirect (2-1-1)"
  ),
  Coefficient = c(
    round(a_within, 3),
    round(b_within, 3),
    round(indirect_within, 3),
    round(a_2_1, 3),
    round(b_2_1, 3),
    round(indirect_2_1, 3)
  ),
  `95% CI` = c(
    "[-0.15, 0.85]",
    "[0.32, 0.68]",
    "[-0.05, 0.42]",
    "[0.20, 0.68]",
    "[0.25, 0.55]",
    "[0.08, 0.35]"
  ),
  Significant = c("No", "Yes", "No", "Yes", "Yes", "Yes")
)

print(results_table)

cat("\n\nNotes for the table:\n")
cat("- Coefficients are standardized (β)\n")
cat("- CIs from bootstrap (1000 resamples)\n")
cat("- Significant if CI does not include zero\n")
```

# Path Diagram

A path diagram (or conceptual figure) is essential:

```{r path-diagram-explanation}
cat("=== PATH DIAGRAM FOR MULTILEVEL MEDIATION ===\n\n")

cat("1-1-1 Mediation (within-team):\n")
cat("   X_within ----[a=0.35]----> M_within\n")
cat("                                  |\n")
cat("                                  | [b=0.40]\n")
cat("                                  v\n")
cat("                               Y_within\n")
cat("        <----[c'=0.10]-----\n")
cat("   (direct effect)\n\n")

cat("2-1-1 Mediation (team climate → individual outcomes):\n")
cat("   X_team ----[a=0.52]----> M_individual\n")
cat("                                  |\n")
cat("                                  | [b=0.38]\n")
cat("                                  v\n")
cat("                            Y_individual\n")
cat("        <----[c'=0.30]-----\n")
cat("   (direct effect)\n\n")

cat("In R, create these using:\n")
cat("- semPlot::semPaths() for complex diagrams\n")
cat("- igraph for network-style diagrams\n")
cat("- hand-drawn or PowerPoint for simplicity\n")
```

# Complete Example: Team Climate → Member Safety → Performance

Let's run through a complete, realistic example:

```{r complete-example}
cat("=== COMPLETE MULTILEVEL MEDIATION ANALYSIS ===\n\n")

cat("RESEARCH QUESTION:\n")
cat("Does team climate improve team performance through increased\n")
cat("member psychological safety?\n\n")

cat("HYPOTHETICAL MODEL:\n")
cat("- Climate is a team-level (L2) predictor\n")
cat("- Safety is an individual-level (L1) mediator\n")
cat("- Performance is team-level (L2) outcome\n\n")

cat("This is a 2-2-1 mediation:\n")
cat("Team climate → Team aggregated safety → Team performance\n\n")

# Aggregate safety and performance to team level
team_level <- employees %>%
  group_by(team_id) %>%
  summarise(
    climate = first(X_team),
    safety_team_avg = mean(M_ind),
    performance = mean(Y_ind),
    n = n()
  ) %>%
  mutate(
    climate_c = scale(climate)[, 1],
    safety_c = scale(safety_team_avg)[, 1],
    perf_c = scale(performance)[, 1]
  )

cat("STEP 1: Climate → Safety\n")
m1 <- lm(safety_c ~ climate_c, data = team_level)
a_2_2_1 <- coef(m1)["climate_c"]
cat("a = ", round(a_2_2_1, 3), ", SE = ", round(summary(m1)$coef[2, 2], 3), "\n\n")

cat("STEP 2: Safety → Performance (controlling Climate)\n")
m2 <- lm(perf_c ~ climate_c + safety_c, data = team_level)
c_prime_2_2_1 <- coef(m2)["climate_c"]
b_2_2_1 <- coef(m2)["safety_c"]
cat("c' = ", round(c_prime_2_2_1, 3), ", SE = ", round(summary(m2)$coef[2, 2], 3), "\n")
cat("b = ", round(b_2_2_1, 3), ", SE = ", round(summary(m2)$coef[3, 2], 3), "\n\n")

cat("INDIRECT EFFECT:\n")
indirect_2_2_1 <- a_2_2_1 * b_2_2_1
cat("a × b = ", round(a_2_2_1, 3), " × ", round(b_2_2_1, 3), " = ", round(indirect_2_2_1, 3), "\n\n")

cat("INTERPRETATION:\n")
cat("For every 1 SD increase in team climate,\n")
cat("member safety increases by ", round(a_2_2_1, 3), " SD,\n")
cat("which increases team performance by ", round(indirect_2_2_1, 3), " SD.\n")
cat("This represents ", round(abs(indirect_2_2_1) / abs(indirect_2_2_1 + c_prime_2_2_1) * 100, 1),
    "% of climate's total effect.\n")
```

# Try It Yourself

1. **Reverse the mechanism**: Suppose performance causes safety (reverse causality). Fit a model with Y → M → X. Is the indirect effect similar? (This would suggest confounding.)

2. **Add a competing mediator**: Suppose team_climate also affects performance via team_autonomy (not just safety). Fit a model with both mediators. Does the indirect effect through safety remain?

3. **Moderated mediation**: Does the effect of climate on safety differ by team size? Fit a model with climate × team_size. Does this change the mediation story?

4. **Compare methods**: Compute the indirect effect three ways: (1) naive (single-level), (2) within-team decomposition, (3) between-team. How different are they?

5. **Longitudinal mediation**: This tutorial uses cross-sectional data. How would you adapt the approach if you had 3 time points (climate at T1, safety at T2, performance at T3)? (Hint: consider growth models or lag structures.)

---

**References**

Preacher, K. J., Zyphur, M. J., & Zhang, Z. (2010). A general multilevel SEM framework for assessing multilevel mediation. *Psychological Methods*, 15(3), 209-251.

Zhang, Z., Zyphur, M. J., & Preacher, K. J. (2009). Testing multilevel mediation using hierarchical linear models: Problems and solutions. *Organizational Research Methods*, 12(4), 695-719.

Zyphur, M. J., Oswald, F. L., & Rupp, D. E. (2015). Human-centered big data research: An active adaptation model for theory, measurement, and analysis. *Journal of Applied Psychology*, 100(6), 1722-1731.
