---
title: "Bayesian Multilevel Modeling and Multilevel Structural Equation Modeling"
subtitle: "PSY 8XXX: Multilevel Modeling for Organizational Research — Week 14"
author: "Instructor Name"
date: last-modified
format:
  html:
    code-fold: true
    code-tools: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

# Introduction

Frequentist multilevel modeling (via `lmer`) has been our workhorse. But **Bayesian approaches** offer distinct advantages in certain scenarios:

1. **Small Level-2 sample sizes**: With few teams or organizations, Bayesian priors can stabilize estimates.
2. **Model complexity**: Non-linear models, complex random effect structures, and missing-data mechanisms fit naturally.
3. **Direct probability statements**: Credible intervals have intuitive interpretation ("given the data and prior, there is 95% probability the parameter lies in this range").
4. **Hierarchical structure as prior**: The multilevel structure itself provides a natural prior for lower-level parameters.

**Structural Equation Modeling (SEM)** at multiple levels adds another dimension: we can model **measurement error** and **latent variables** that vary across levels. For instance, we might estimate a latent "team climate" at Level 2, or individual-level latent factors with cross-level effects.

This week, we build intuition for Bayesian MLM and introduce multilevel CFA.

---

# Bayesian Estimation: The Core Idea

Bayesian inference combines three components:

$$\text{Posterior} \propto \text{Likelihood} \times \text{Prior}$$

**Intuitively:**
- **Prior**: our belief about parameters *before* seeing data
- **Likelihood**: how likely the data are, given parameters
- **Posterior**: our updated belief *after* observing data

For multilevel data:
- The **prior for lower-level parameters** often comes from the *higher-level distribution*
- E.g., the prior on each team's intercept is the organization-level distribution
- This is automatic in Bayesian MLM

**Why this matters in MLM:**
- With 5 teams, we have sparse information for estimating team-level variance. Bayesian methods, properly specified, can pool information intelligently.
- With 500 teams, frequentist and Bayesian approaches converge (priors matter less when data are abundant).

---

# Setting Up brms

The `brms` package ("Bayesian Regression Models using Stan") provides a formula interface similar to `lmer`, but with Bayesian inference:

```{r}
# Install if needed: install.packages("brms")
library(tidyverse)
library(lme4)
library(lmerTest)
library(brms)
library(tidybayes)
library(bayesplot)

# Set seed for reproducibility
set.seed(1234)
```

We'll use the same data structure as in earlier weeks. Let me create a simplified three-level dataset:

```{r}
# Simulate data: employees in teams in organizations
n_orgs <- 8
n_teams_per_org <- 6
n_per_team <- 12

n_teams <- n_orgs * n_teams_per_org
n_total <- n_teams * n_per_team

# Generate data
org_data <- tibble(
  org_id = 1:n_orgs,
  org_climate = rnorm(n_orgs, mean = 5, sd = 1)
)

team_data <- expand_grid(
  org_id = 1:n_orgs,
  team_num = 1:n_teams_per_org
) %>%
  mutate(
    team_id = row_number(),
    team_safety = rnorm(n_teams, mean = 5, sd = 0.8)
  ) %>%
  left_join(org_data, by = "org_id")

employee_data <- expand_grid(
  team_id = 1:n_teams,
  emp_num = 1:n_per_team
) %>%
  mutate(
    person_id = row_number(),
    effort = rnorm(n_total, mean = 5, sd = 1.5)
  ) %>%
  left_join(team_data, by = "team_id") %>%
  mutate(
    # Outcome: performance
    # Depends on effort, team safety, org climate
    performance = 5 +
      0.35 * scale(effort)[,1] +
      0.25 * scale(team_safety)[,1] +
      0.20 * scale(org_climate)[,1] +
      rnorm(n_total, mean = 0, sd = 0.8)
  ) %>%
  select(person_id, team_id, org_id, effort, team_safety, org_climate, performance)

# Quick summary
head(employee_data, 12)
```

---

# Fitting a Bayesian Multilevel Model

The `brms` syntax mirrors `lmer`, with formula specification via `bf()`:

```{r, eval = FALSE}
# Bayesian two-level model with brms
# Note: eval=FALSE because brms requires Stan compilation, which takes time
model_bayes_l2 <- brm(
  formula = bf(performance ~ effort + team_safety + org_climate + (1 | team_id)),
  data = employee_data,
  family = gaussian(),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 4,
  seed = 1234,
  verbose = FALSE
)

summary(model_bayes_l2)
```

**Key parameters:**
- **chains**: number of MCMC chains (typically 4, samples in parallel)
- **iter**: total iterations per chain
- **warmup**: "burn-in" iterations discarded before inference
- **cores**: parallel processing

When you run this locally, Stan will compile and sample from the posterior. The output resembles frequentist models but with additional diagnostics.

**Note**: This tutorial sets `eval=FALSE` for brms chunks since Stan compilation requires significant computation. Run these on your own machine.

---

# Specifying Priors

By default, `brms` uses weak/non-informative priors. But we can specify priors directly:

```{r, eval = FALSE}
# Check default priors
default_priors <- prior_summary(model_bayes_l2)

# Specify custom priors
priors_custom <- c(
  # Priors on fixed effects
  prior(normal(5, 2), class = "b", coef = "effort"),           # effort effect
  prior(normal(0, 1), class = "b", coef = "team_safety"),      # team safety effect
  prior(normal(0, 1), class = "b", coef = "org_climate"),      # org climate effect
  # Prior on intercept
  prior(normal(5, 1), class = "Intercept"),
  # Prior on standard deviations
  prior(exponential(1), class = "sigma"),                       # residual SD
  prior(exponential(1), class = "sd")                           # random effect SDs
)

# Fit with custom priors
model_bayes_priors <- brm(
  formula = bf(performance ~ effort + team_safety + org_climate + (1 | team_id)),
  data = employee_data,
  prior = priors_custom,
  family = gaussian(),
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 4,
  seed = 1234
)

summary(model_bayes_priors)
```

**Prior choices:**

- **`normal(5, 2)` on effort**: we *weakly expect* the effort effect to be around 0.3–0.4 (on a scale of 1–10), but allow wide uncertainty
- **`normal(0, 1)` on other effects**: centered at 0, allowing both positive and negative effects
- **`exponential(1)` on SDs**: ensures positive standard deviations, with most prior mass on moderate values

These are **weakly informative priors**: they don't dominate the likelihood but express reasonable domain knowledge.

---

# Interpreting Bayesian Output

Key differences from frequentist output:

```{r, eval = FALSE}
# Extract posterior samples
posterior_samples <- as_draws_df(model_bayes_l2)

# Posterior means and credible intervals
posterior_summary(model_bayes_l2)
```

**Output includes:**

- **Estimate**: posterior mean
- **Est.Error**: posterior standard deviation
- **Q2.5, Q97.5**: 95% credible interval (quantiles of posterior)
- **Rhat**: diagnostic convergence statistic (should be < 1.01)
- **Bulk_ESS / Tail_ESS**: effective sample size (higher is better)

**Interpretation:**

- A **credible interval** for the effort effect of [0.30, 0.40] means: "Given the data and prior, we believe the true effect lies in this range with 95% probability."
- Compare to frequentist confidence interval: "If we repeated the study many times, 95% of confidence intervals would contain the true parameter" (subtly different framing).

---

# Visualizing Posterior Distributions

One advantage of Bayesian methods is rich posterior visualization:

```{r, eval = FALSE}
# Posterior plots
library(bayesplot)

# Plot posterior distributions for fixed effects
plot(model_bayes_l2)

# More detailed: extract and plot manually
posterior_samples %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(x = value, fill = name)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~name, scales = "free") +
  labs(title = "Posterior Distributions of Fixed Effects",
       x = "Parameter Value",
       y = "Posterior Density") +
  theme_minimal()

# Posterior predictive checks: do samples from posterior match observed data?
pp_check(model_bayes_l2)  # overlay replicated data on observed
```

These visualizations help assess:
1. **Convergence**: are chains mixing well?
2. **Posterior predictive validity**: do the model's predictions match the observed data?

---

# Model Comparison: LOO and WAIC

Bayesian model comparison uses **Leave-One-Out Cross-Validation (LOO)** or **WAIC** (Widely Applicable Information Criterion):

```{r, eval = FALSE}
# Add LOO criterion to models
model_bayes_l2 <- add_criterion(model_bayes_l2, "loo")
model_bayes_priors <- add_criterion(model_bayes_priors, "loo")

# Compare models
loo_compare(model_bayes_l2, model_bayes_priors)
```

**Interpretation:**

- **elpd_diff**: difference in expected log predictive density (positive favors first model)
- **se_diff**: standard error of difference (if |elpd_diff| > 2*se_diff, meaningful difference)

This is conceptually similar to AIC/BIC, but has theoretical advantages in Bayesian frameworks.

---

# When Bayesian Helps: Example with Small L2 Sample

To illustrate Bayesian advantages, let's compare frequentist and Bayesian estimates with **only 5 teams**:

```{r}
# Subsample to 5 teams only
set.seed(1234)
small_sample <- employee_data %>%
  filter(team_id <= 5)

# Frequentist model
model_freq_small <- lmer(
  performance ~ scale(effort)[,1] + (1 | team_id),
  data = small_sample
)

# Display variance estimates
cat("Frequentist variance estimate (team SD):\n")
print(VarCorr(model_freq_small))
```

With only 5 teams, the frequentist variance estimate is based on sparse information. Bayesian priors can stabilize this:

```{r, eval = FALSE}
# Bayesian model with informative prior on team variance
priors_small <- c(
  prior(normal(5, 2), class = "Intercept"),
  prior(normal(0, 1), class = "b"),
  prior(exponential(1), class = "sigma"),
  prior(exponential(1), class = "sd")  # prior on team SD
)

model_bayes_small <- brm(
  formula = bf(performance ~ scale(effort)[,1] + (1 | team_id)),
  data = small_sample,
  prior = priors_small,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  cores = 4,
  seed = 1234
)

# Bayesian variance estimate shrinks toward prior, stabilizing inference
posterior_summary(model_bayes_small)
```

The key insight: **Bayesian estimates trade some bias for lower variance**, which can improve prediction and reduce false discoveries when L2 samples are small.

---

# Introduction to Multilevel Structural Equation Modeling

So far, we've assumed predictors and outcomes are **observed without error**. But in organizational research, many constructs are **latent**: we measure them via multiple indicators (e.g., a 5-item engagement scale).

**Multilevel SEM** extends multilevel models to include:
1. **Latent variables** at each level
2. **Measurement models** (factor structure)
3. **Structural models** (relationships between latents)

Example research question: "How does team climate (latent construct, measured via 4 items) predict individual engagement (latent, measured via 3 items) after accounting for measurement error at both levels?"

---

# Multilevel CFA with lavaan

`lavaan` is R's primary SEM package. Multilevel CFA is specified via `level: 1` and `level: 2` in model definition.

## Simulating Multilevel Factor Data

```{r}
# Simulate data with latent team climate and individual engagement
set.seed(1234)

n_teams_cfa <- 20
n_per_team_cfa <- 25
n_total_cfa <- n_teams_cfa * n_per_team_cfa

# Latent true scores and indicators
team_climate_true <- rep(rnorm(n_teams_cfa, mean = 5, sd = 1), each = n_per_team_cfa)
engagement_true <- rnorm(n_total_cfa, mean = 5.5, sd = 1.5) +
                   0.4 * team_climate_true +  # team climate predicts engagement
                   rep(rnorm(n_teams_cfa, 0, 0.5), each = n_per_team_cfa)

# Generate measured indicators (with error)
# Team Climate: 4 indicators
tc_ind1 <- 0.8 * team_climate_true + rnorm(n_total_cfa, 0, 0.6)
tc_ind2 <- 0.85 * team_climate_true + rnorm(n_total_cfa, 0, 0.5)
tc_ind3 <- 0.75 * team_climate_true + rnorm(n_total_cfa, 0, 0.7)
tc_ind4 <- 0.82 * team_climate_true + rnorm(n_total_cfa, 0, 0.6)

# Engagement: 3 indicators
eng_ind1 <- 0.8 * engagement_true + rnorm(n_total_cfa, 0, 0.6)
eng_ind2 <- 0.85 * engagement_true + rnorm(n_total_cfa, 0, 0.5)
eng_ind3 <- 0.78 * engagement_true + rnorm(n_total_cfa, 0, 0.65)

cfa_data <- tibble(
  team_id = rep(1:n_teams_cfa, each = n_per_team_cfa),
  tc1 = tc_ind1, tc2 = tc_ind2, tc3 = tc_ind3, tc4 = tc_ind4,
  eng1 = eng_ind1, eng2 = eng_ind2, eng3 = eng_ind3
)

head(cfa_data, 8)
```

## Specifying Multilevel CFA in lavaan

```{r, eval = FALSE}
library(lavaan)

# Define multilevel CFA model
# level: 1 specifies within-level structure (individual variation)
# level: 2 specifies between-level structure (team variation)

mlm_cfa_model <- "
  level: 1
    climate =~ tc1 + tc2 + tc3 + tc4
    engagement =~ eng1 + eng2 + eng3
    engagement ~ climate

  level: 2
    climate =~ tc1 + tc2 + tc3 + tc4
    engagement =~ eng1 + eng2 + eng3
    engagement ~ climate
"

# Fit the model
mlm_cfa_fit <- cfa(mlm_cfa_model,
                    data = cfa_data,
                    cluster = "team_id")

# Summary
summary(mlm_cfa_fit, fit.measures = TRUE)
```

**Output interpretation:**

The model estimates:
- **Within-level (L1)**: individual loadings, factor variances, and the climate → engagement path
- **Between-level (L2)**: team-level latent factor structure and the team climate → team engagement relationship

This cleanly partitions measurement error at each level, crucial when constructs are measured at different granularities.

## Practical Multilevel CFA Example

A realistic scenario: employee engagement survey in a 25-team organization.

```{r, eval = FALSE}
# Simplified two-factor model: autonomy and support (both predictors of engagement)
mlm_model_expanded <- "
  level: 1
    autonomy =~ aut1 + aut2 + aut3
    support =~ sup1 + sup2 + sup3
    engagement =~ eng1 + eng2 + eng3
    engagement ~ autonomy + support

  level: 2
    autonomy =~ aut1 + aut2 + aut3
    support =~ sup1 + sup2 + sup3
    engagement =~ eng1 + eng2 + eng3
    engagement ~ autonomy + support
"

# Notes:
# - At L1, we estimate how individual autonomy/support predict individual engagement
# - At L2, we estimate how team-average autonomy/support predict team-average engagement
# - Measurement models are estimated at both levels, accounting for indicator error
```

The elegance of multilevel CFA: we simultaneously account for measurement error and multilevel structure, answering: "How does team context (measured via indicators with error) affect individual outcomes (also measured with error)?"

---

# Practical Guidance

## When to Use Bayesian MLM
- **Small L2 samples** (< 15 clusters)
- **Complex models** (multiple random slopes, non-linear terms)
- **Strong domain knowledge** to specify priors
- **Direct probability statements** are desired

## When to Use Multilevel CFA
- **Latent variables** are theoretically important
- **Measurement error** is substantial
- **Different levels** have different constructs
- **Published scales** with multi-item indicators

## Computational Considerations
- **brms** is powerful but computationally intensive (Stan sampling takes time)
- **lavaan** is relatively fast for CFA and SEM
- Start simple: fit one-level models before moving to multilevel
- Use informative priors in Bayesian models to aid convergence

---

# Try It Yourself

## Exercise 1: Prior Sensitivity

Fit a Bayesian model with **three different prior specifications**:
1. Non-informative (very weak priors)
2. Weakly informative (matching domain knowledge)
3. Informative (tight priors on expected values)

Compare posterior estimates and credible intervals. How much do priors influence conclusions when L1 sample size is large? When small?

## Exercise 2: Posterior Predictive Checks

For a fit Bayesian model, use `pp_check()` to examine:
- Do replicated datasets from the posterior match the observed data?
- Are there systematic discrepancies?
- Visually inspect and describe findings.

## Exercise 3: Multilevel CFA Model Extension

Extend the multilevel CFA example to include:
- Three latent factors at L1 (e.g., autonomy, support, opportunity for development)
- All three predicting a final engagement factor
- Estimate the model and interpret path coefficients at L1 and L2

## Exercise 4: Hybrid Bayesian-Frequentist Workflow

- Fit a frequentist MLM (lmer) and extract estimates
- Specify Bayesian priors *centered on* frequentist point estimates
- Fit a Bayesian model with these priors
- Compare inference: do results converge or diverge? Explain any differences.

